---
title: "Welcome to Introduction to R"
description: |
  Welcome to the course page for Introduction to R!
author:
  - name: Garrick Aden-Buie 
    url: https://www.garrickadenbuie.com
    affiliation: Collaborative Data Services, Moffitt Cancer Center
    affiliation_url: https://moffitt.org
date: 10-23-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Learn more about creating blogs with Distill at:
# https://rstudio.github.io/distill/blog.html

```


Learn more about creating blogs with Distill at:
<https://rstudio.github.io/distill/blog.html>.

Create a new post by running

```{r eval=FALSE}
library(distill)

create_post("Topic for Week X", draft = TRUE)
```

When your topic post is complete, remove the `draft: true` from the post's YAML metadata.

## Install ggplot2

We're going to work with the ggplot2 package. Install and load it:

```{r eval=FALSE}
install.packages("ggplot2")
```

```{r}
library(ggplot2)
```

## Our data

```{r echo=FALSE}
write.csv(mtcars, "mtcars.csv")
```

Download the data for this lesson [here](mtcars.csv).

## Slides

Here are [the slides for today's lesson](../../slides/00_slide-starter.html).

## Scripts For Today's Lesson

Here's an R Script for our lesson: [plumber.R](plumber-demo-script.R).