---
title: "ggplot2 exercise"
author: "Ram Thapa"
date:  "April 21, 2021"
output: 
  html_document:
    df_print: paged
    code_folding: show
    toc: true
    toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 4, fig.width = 5, fig.align = "center", warning = FALSE)
```

Data visualization is important in most phases of data analysis workflow and being able to create visualizations of data is a key step in being able to effectively communicate information and findings to others. 

{ggplot2} is a plotting framework that is (relatively) easy to use, powerful and hassle-free. It is based on a set of principles "Grammar of graphic" and is designed to work in a layered fashion, starting with a layer showing the data then adding layers of graphical objects, annotations and statistical summaries.

**Packages and data**

```{r packages, message = FALSE, warning = FALSE}
library(tidyverse)
library(patchwork)
library(ggridges)

pancancer <- readRDS(here::here("slides", "data", "pancancer.RDS"))

head(pancancer)
```

## Visualizing associations among two or more variables

### Scatterplot

Relationship between height vs weight

```{r scatterplot}
# set theme to theme_bw()
theme_set(theme_bw(base_size = 15))

wt_ht_scatter <- ggplot(
  data = pancancer,
  mapping = aes(x = weight, y = height)) +
  geom_point()

wt_ht_scatter
```

Mapping `color` to `gender` variable from the data

```{r scatterplot2, fig.width = 7}
wt_ht_scatter_gender <- ggplot(
  data = pancancer,
  mapping = aes(x = weight, y = height, color = gender)) +
  geom_point()

wt_ht_scatter_gender
```

Avoid overplotting using jitter plot

```{r jitter,  fig.width = 7}
pancancer %>% 
  drop_na(stage_collapsed) %>% 
  ggplot(mapping = aes(x = stage_collapsed, y = height)) +
  geom_point() +
  geom_jitter(aes(color = stage_collapsed), width = 0.2, size = 2, alpha = 0.7) +
  labs(x = NULL, y = "Height (cm)") +
  scale_color_brewer(NULL, palette = "Paired") 
```

Note: `geom_jitter` is a shortcut for `geom_point(position = "jitter")`

### Line plot

Line plot of mean age by Karnofsky Performance Score for each stage

```{r line, fig.width = 7}
# mean age
mean_age_kps <- pancancer %>% 
  drop_na(stage_collapsed, karnofsky_performance_score) %>% 
  group_by(stage_collapsed, karnofsky_performance_score) %>% 
  summarise(mean = mean(age_at_initial_pathologic_diagnosis, na.rm = TRUE))

# plot
mean_age_kps_line <- mean_age_kps %>% 
  ggplot(aes(x = factor(karnofsky_performance_score), y = mean, 
             group = stage_collapsed, color = stage_collapsed)) +
  # geom_point(size = 4) +
  geom_line() +
  scale_x_discrete("Karnofsky Performance Score") +
  scale_y_continuous("Mean age at diagnosis") +
  scale_color_brewer("Stage", palette = "Paired") 

mean_age_kps_line
```


## Visualizing amounts

We might be interested in visualizing amounts i.e. numerical values for some set of categories

### Bar plot

```{r barplot}
pancancer %>% 
  ggplot(aes(x = stage_collapsed)) +
  geom_bar(fill = "gray60")
```

You could also calculate counts in advance and plot it but you need to use `stat = "identity"`

```{r count}
# Count by stage
stage_count <- pancancer %>% 
  count(stage_collapsed, .drop = TRUE)

# use stat = "identity"
ggplot(data = stage_count, aes(x = stage_collapsed, y = n)) +
  geom_bar(stat = "identity", fill = "gray60") +
  labs(x = "Stage", y = "Number of patients")
```

Or, just use `geom_col()`

```{r count2}
ggplot(data = stage_count, aes(x = stage_collapsed, y = n)) +
  geom_col(fill = "gray60") +
  labs(x = "Stage", y = "Number of patients")
```

Sometimes x-axis labels identifying each bar take up a lot of horizontal space

```{r}
drop_na(stage_count, stage_collapsed) %>% 
  ggplot(
    aes(y = fct_reorder(stage_collapsed, n), x = n)) +
  geom_bar(stat = "identity", fill = "gray60") +
  labs(x = "Stage", y = NULL) 
```

Grouped bar plot

```{r grouped_bar, fig.width = 7}
groupedbar_gender <- pancancer %>% 
  drop_na(stage_collapsed) %>% 
  ggplot(aes(x = stage_collapsed, fill = gender)) +
  geom_bar(position = "dodge") +
  labs(x = NULL, y = "Number of patients") +
  scale_fill_viridis_d(NULL)

groupedbar_gender
```

Stacked bar plot

```{r stacked_bar, fig.width = 6}
stackedbar_gender <- pancancer %>% 
  drop_na(stage_collapsed) %>% 
  ggplot(aes(x = stage_collapsed, fill = gender)) +
  geom_bar(position = "stack") +
  labs(x = NULL, y = "Number of patients") +
  scale_fill_viridis_d(NULL)

stackedbar_gender
```

## Visualizing distributions

To understand how a particular variable is distributed in a dataset

### Boxplot

```{r boxplot}
ggplot(
  data = pancancer, 
  aes(x = gender, y = weight, fill = gender)) +
  geom_boxplot(show.legend = FALSE) +
  labs(x = NULL, y = "Weight (kg)") +
  scale_fill_brewer(palette = "Dark2")
```

### Violin plot

Similar to box plots except they also show the kernel probability density of the data at different values

```{r violin, fig.width = 8}
pancancer %>% 
  drop_na(stage_collapsed) %>% 
  ggplot(
    aes(x = stage_collapsed, y = weight, fill = gender)) +
  geom_boxplot() +
  labs(x = NULL, y = "Weight (kg)") +
  scale_fill_brewer(NULL, palette = "Dark2")

```


### histogram and density plot

Histogram count plot for single distribution

```{r histogram}
hist_age <- ggplot(pancancer, aes(x = age_at_initial_pathologic_diagnosis)) +
  geom_histogram(fill = "gray60", color = "gray80") +
  labs(x = "Age at diagnosis")

hist_age
```

Density plot for visualizing multiple distributions at the same time

```{r density, fig.width = 7}
ggplot(
  pancancer, 
  aes(x = height, fill = gender)) +
  geom_density(alpha = 0.8, color = "transparent") +
  labs(x = "Height (cm)") +
  scale_fill_viridis_d(NULL) 
```

### Ridge plot

Visualizing multiple distributions along the horizontal axis by staggering the distribution plots in the vertical direction

You need `{ggridges}` package for this plot

```{r ridge, fig.width = 7}
ridge_height <- pancancer %>% 
  drop_na(stage_collapsed) %>% 
  ggplot(
  aes(x = height, y = fct_rev(factor(stage_collapsed)))) +
  geom_density_ridges(
    scale = 1.5, fill = "skyblue", color = "white") +
  geom_density(alpha = 0.8, color = "transparent") +
  labs(x = "Height (cm)", y = NULL) +
  scale_fill_viridis_d(NULL) 

ridge_height
```

## Combining plots

`{patchwork}` R package has made combining multiple plots ridiculuous simple

### Horizontal layout

```{r, fig.width = 8}
# combine two plots horizontally
wt_ht_scatter + mean_age_kps_line
```

or you can use "|" too

### Vertical layout

```{r, message = FALSE, fig.height = 6}
# stack two plots
ridge_height / hist_age
```

### Nested layout

```{r, message = FALSE, fig.height = 6, fig.width = 8}
patch <- mean_age_kps_line / (wt_ht_scatter | hist_age)

patch
```

Be careful with the operator precedence rule (e.g. / is evaluated before |)

`{patchwork}` provides `wrap_plots()` for a more functional approach to assembly

### Annotating plots

```{r, message = FALSE, fig.height = 6, fig.width = 8}
(ridge_height / (wt_ht_scatter | hist_age)) +
  plot_annotation(
    title = "Combining three plots",
    subtitle = "Annotating each plot",
    caption = "Data source: TCGA data",
    tag_levels = "A", tag_suffix = ".")
```   

### Modifying patches

Use `&` to add elements to all subplots in the patchwork

```{r, message = FALSE, fig.height = 6, fig.width = 8}
mean_age_kps_line / (wt_ht_scatter | hist_age) & theme_gray()
```

and `*` to add the element to all the subplots in the current nesting level

```{r, message = FALSE, fig.height = 6, fig.width = 8}
(mean_age_kps_line / (wt_ht_scatter | hist_age)) * theme_gray()
```

### Controlling guides

If multiple plots have same guide, using `guides = "collect"` in `plot_layout()` drops the duplicate guide

```{r, message = FALSE, fig.height = 6, fig.width = 8}
(groupedbar_gender | (stackedbar_gender / hist_age)) +
  plot_layout(guides = "collect")
```
