---
title: "Data warangling : part I (answers)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

# 5 basic verbs. also mutate_all, mutate_if, summarise_all, summarise_if
# select: subset columns. comes with helper functions like starts_with, ends_width, contains, etc.
# filter: subset rows
# summarise: aggregates - british spelling
# mutate: add column(s) to existing data
# arrange: orders data "desc"
```

## Importing TCAG data 

Import both TCGA data files using either the "Import Dataset" GUI or by writing
commands. Then take a look at the data. 

```{r}
library(readr)
clinical <- read_csv(here::here("data/tcga-clinical.csv"))
gene_exp <- read_csv(here::here("data/tcga-gene-exp.csv"))

# str(clinical)
# glimpse(clinical)
# head(clinical)
# View(clinical)
```

## select()

Show the first 4 columns of the clinical data (hint there are multiple ways to do this)

```{r}
clinical[, 1:4] # traditional method
clinical[, c("bcr_patient_barcode", "acronym", "tobacco_smoking_history", "race")] # traditional method
select(clinical, bcr_patient_barcode, acronym, tobacco_smoking_history, race) # no quotes around variable names
select(clinical, bcr_patient_barcode:race)
select(clinical, 1:4)
select(clinical,-(5:14))

```

## filter()

Filter clinical data to individuals who have kidney renal cell carcinoma (KIRC) 
and were diagnosed over the age of 60.

```{r}
clinical[clinical$acronym == "KIRC" & clinical$age_at_initial_pathologic_diagnosis > 60,] # traditional method
filter(clinical , acronym == "KIRC", age_at_initial_pathologic_diagnosis > 60)
filter(clinical, acronym == "KIRC" & age_at_initial_pathologic_diagnosis > 60)
```

## summarize()

```{r}

mean(clinical$age_at_initial_pathologic_diagnosis, na.rm = TRUE) # traditional method
summarise(clinical, mean(age_at_initial_pathologic_diagnosis, na.rm = TRUE))
summarise(clinical, mean_age = mean(age_at_initial_pathologic_diagnosis, na.rm = TRUE))
summarise(clinical, mean_age = mean(age_at_initial_pathologic_diagnosis, na.rm = TRUE),
                    total_patients = n(),
                    unique_cancers = n_distinct(acronym)
)
```

## mutate()

Create a new variable of the log of PTEN expression and call it pten_log

```{r}

```

## arrange()

Order gene expression data by DNAH12 expression then DNAH10 expression

```{r}

```

## piping

Piping allows you to combine steps - "and then" and avoids nesting functions or needing
to create new data frames for each manipulation 

```{r}

```

## group_by()

Allows you to apply the "verbs" on the object "by group"

```{r}

```

## case_when()

Replaces the traditional `ifelse()`

```{r}


```
