---
title: "My super cool document"
author: 
  - Jordan Creed
output: html_document
---

```{r setup}
library(tidyverse)

knitr::opts_chunk$set(echo = TRUE)
```

Formatted text:

-   *italics*
-   **bold**
-   `code`
-   [links](https://www.tidyverse.org/)

```{r}
# example code chunk 

x <- seq(0,10,2)
x
```

Sequential lines will be
rendered as
a
 single line.
 
### Activity 1

Write a few sentences introducing yourself and add a picture of your favorite
animal below.

Then re-render your document. 

### Activity 2

Add a parameter to this report changing a few elements in the sentences you
wrote for activity one. This should make it so that you just need to 
change parameters (hint found in YAML) in order to create a new introduction
paragraph. 

### Activity 3

The are of a circle is equal to $\pi r^{2}$. Write a code chunk that calculates
the area of circles with radi of 748, 35, 2.986, and write a summary sentence,
with the answers included. 

### Activity 4

Change the output this document to output a PDF document and word document. 

