---
title: "06-final-workflow"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


## Loading and cleaning data 

Import the files `CIplus-cases.txt` and `CIplus-population.txt` and combine 
them into a single dataframe called ciplus. Create a new variable for the proportion (%) of cases in the specific population (not total population). 

Using the tidyverse, which cancers are most common in men and women 
who are aged 30-39 in the year 2012? 

**Hint:** You may need to use multiple `group_by` statements and a `slice`
statement to take a specific value.

```{r}

```

## Plotting

Plot total number of cases by total population for all cancers 
(excluding the All cancers excluding non-melanoma skin group) in 2000 in Non-Hispanic White subjects - calculate total counts first. 

Use a different color for each cancer type. 

**Hint:** You will need to following line to clean up cancer names: `mutate(cancer = str_extract(cancer, "^.*(?=:)"))`

```{r}

```

## Analysis

For all cancers excluding non-melanoma skin in 1992 in Hispanic-White subjects was there a statistical association between gender and cancer cases?
Use fishers exact test or a chi-square test.

```{r}

```



